\chapter{Algorithms}

\section{Sorting}
\subsection{Selection sort}
\paragraph{Algorithm Description.} Sort \(n\) numbers stored in array
\(\mathbf{A}\) by finding the smallest element of \(\mathbf{A}\) and exchanging
it with the element in \(\mathbf{A}[1]\). Then find the second smallest element
of \(\mathbf{A}\), and exchange it with \(\mathbf{A}[2]\). Continue in this
manner for the first \(n - 1\) elements of \(\mathbf{A}\).

\textbf{Implementation.}

\begin{lstlisting}
for i = 1 to A.length - 1
  // sI stands for smallestIndex
  sI = i
  for j = i + 1 to A.length
    if A[j] < A[sI]
      sI = j
  // swap (if necessary)
  if (sI != i)
    tmp = A[i]
    A[i] = A[sI]
    A[sI] = tmp
\end{lstlisting}

\textbf{Analysis.} The best-case for this algorithm is an already sorted data
array. The worst-case would be produced by a sorted array with the first element
moved to the last position and all the other elements moved one place to the
left. Note that the only thing that varies is the amount of swaps made, the
algorithm has \(\Theta(n^2)\) time-complexity even for sorted input.

\section{String matching}
\subsection{Rabin-Karp (1987)}
This algorithm speeds up equality testing by using a hash function.

If the hash of the substring from the text was recomputed at each iteration,
this would become an O(mn)-time algorithm, this problem is solved by reusing
the last computed hash in the evaluation of the new hash. Such a hash function
is called a \textbf{rolling hash}.

\textbf{Rabin Fingerprint}. This hash treats every substring as a number in a
base \(\alpha\), usually a large prime.

\begin{lstlisting}
  // A Rabin Fingerprint roll.
  hash -= first_char * a ^ (m - 1)
  hash *= a
  hash += new_last_char
\end{lstlisting}

This algorithm is slower than Knuth-Morris-Pratt and Boyer-Moore for single
pattern string searching because of its slow worst case behavior. However, it is
an algorithm of choice for multiple pattern search.

For instance, to find if any string of a large number of strings is in a text we
can create a variant of the Rabin-Karp algorithm that uses a set data structure
to check whether the hash of a substring belongs to a set of hash values of
patterns in constant time.
